---
title: Intro to `data.table`
fontsize: 9pt
output:
  beamer_presentation:
    theme: "Dresden"
    colortheme: "rose"
    fonttheme: "professionalfonts"
---

# Why are we here?

You want to know about `data.table`.

I know somethings about `data.table`.

Why do I use it?

Let's load `data.table` and find out!

```{r}
library(data.table)
```

# It's terse --->>> *It's fast* --->>> It's `base`-ic

`data.table` is a short jump from base R and you get a big boost in speed with less typing.

Reading in data and subsetting is really familiar.

```{r}
# base R
df <- read.csv("./data/slides/mergtab.csv")
df[df$fname== "Jack",][1:2,]

# data.table
dt <- fread("./data/slides/mergtab.csv")
dt[fname == "Jack",][1:2,]
```

`fread()` is fast, versatile, and never defaults strings to factors.

# `dt[i, j, by / with / on / SDCols]`

Base R subsetting using `[` already had a SQL like interface.

This...

```{r, eval = F}
df[<subset rows expression> , <subset columns expression>]
```

...is like saying...

```{r, eval = F}
df[<WHERE> , <SELECT>]
```

# `dt[i, j, with / by / on / SDCols]`

`data.table`'s `[` function acts like base R but has some additional arguments.

`[i, j]` work the same, more or less, like SQL:

`dt[<WHERE>, <SELECT>]`

* `with` allows to call columns using strings
* `by` allows for performing operations in groups of unique values, like SQL's `GROUP BY` operation
* `on` is used when using `[` as a merge operator

SPOILER ALERT: `j` can evaluate whole expressions! 

# Using `with` and `data.table`'s variable scope

Unlike `data.frame`, `data.table` will look for variables using the `data.table` object's namespace.

What does this mean? Less typing!

```{r}
# base R
df[df$fname == "Jack" &
   as.Date(df$date) > as.Date("2019-11-02") &
   df$type == "desktop",]
unique(df[, c("lname")])
```

# Using `with` and `data.table`'s variable scope

```{r}
# data.table
dt[fname == "Jack" & date > "2019-11-02" & type == "desktop",]
unique(dt[, .(fname, lname)])
```

# Using `with` and `data.table`'s variable scope

In base R, you can use `with()` to search a data.frame's scope for variable names.

```{r}
with(df, unique(uid))
with(df, df[fname == "Jill" & type == "phone",])
```

# Using `with` and `data.table`'s variable scope

`data.table` does this by default.

```{r}
unique(dt[,uid])
dt[fname == "Jill" & type == "phone",]
```

# Using `with` and `data.table`'s variable scope

With data.frame you can pass a variable of column names to subset a table.

```{r}
colNames <- c("fname", "type", "date")
df[df$fname == "Gretel", colNames][1,]
```

Using `with = FALSE` brings back that behavior to data.table, that is, if you store the column names in a variable then use `with = FALSE` to call those names as you would with a `data.frame`.

```{r}
colNames <- c("fname", "type", "date")
dt[fname == "Gretel", colNames, with = F][1,]
```

# Using `with` and `data.table`'s variable scope

This main message here is that a `data.table` object will always look
in it's own namespace first for variables. This makes subsetting a
`data.table` object easier.

# Using `by`

`by` acts like SQL `GROUP BY`. It performs operations by unique values in a given column on an expression passed to `j`.

```{r}
dt[, sum(type == "desktop"), by = .(fname)]
```

You could also do:

```{r, eval = F}
dt[, sum(type == "desktop"), by = c("fname")]
```

# Using `by` with `.()`

Now is also a good time to introduce `.()`. 

`.()` is `data.table` short hand for `list()` and it is used for concatenating variables in a `data.table`'s namespace

```{r}
unique(dt[, .(fname, type)])
```

# Using `by` with `.()`

```{r}
dt[, .(session_cnt = sum(type == "phone")), by = .(fname)]
```

# Chaining `by`

Chaining `data.table` is awesome. 

```{r}
dt[, .(con_cnt = sum(type %like% "desktop|phone")), by = .(fname)
   ][, .(fname, con_cnt, con_perc = con_cnt / sum(con_cnt) * 100)]
```

You can chain `data.table` all day long...

```{r}
dt[, .(con_cnt = sum(type %like% "desktop|phone")), by = .(fname)
   ][, .(fname, con_cnt, con_perc = con_cnt / sum(con_cnt) * 100)
     ][con_perc == max(con_perc),]
```

# Adding columns by reference

`data.table` allows for adding columns by reference and uses an
operator type syntax, `:=`.

This makes a big difference in performance when working with large
datasets.

```{r}
dt[, isGerman := ifelse(ccode == "DE", 1, 0)]
dt[, isGermanPhone := as.numeric(isGerman & type == "phone")]
head(dt[,-c("date")])
```

# Adding columns by reference

And, this operation can be done multiple times as a single call using
two different methods.

```{r}
## calling `:=` using function call syntax
dt[, `:=`(
    isGerman = ifelse(ccode == "DE", 1, 0),
    isGermanPhone = as.numeric(isGerman & type == "phone")
    )]

## calling `:=` through chaining
dt[, isGerman := ifelse(ccode == "DE", 1, 0)
   ][, isGermanPhone := as.numeric(isGerman & type == "phone")]

```

# Using `.N` and `.SD`

`data.table` has special variables that can be used in `j`.

* `.N` counts the number of records in a given group
* `.SD` passes a subset of a group's `data.table`

`.N` adds a row count as a field. This is similar to `table()`.

```{r}
dt[, .N, by = fname]
```

# Using `.N`

```{r}
dt[, .N]
dt[, .N, by = .(fname, type)
   ][N > 3 & type == "phone"]
```

# Using `.SD` - "*S*ub *D*ata table"

This will return subsets of the data.table object using the `by`
argument. The use of `.SD` can best be shown using a print statement
in `j`.

```{r}
dt[,
   print(
       .SD[, .N, by = .(fname, lname)]
   ), by = .(uid)]
```

# Using `.SD`

```{r}
dt[,
   .SD[, .(
       maxDate = max(date),
       totalPhone = sum(type == "phone"),
       totalDesktop = sum(type == "desktop")
   )],
   by = .(uid)]
```

# Using `merge`

`merge` can be used in two ways:

* the base-ic way: `merge(x, y)`
* the data.table way: `y[x]`

That's right, `[` is also used for merges. 

# Using `[` for merging

I go back and forth using this syntax. 

`[` feels a bit too implicit to me, but it's great in chains.

`y[x]`

This is equivalent to `merge(x, y, all.x = T)`

Using this syntax, you also need to use the `on` argument.

`y[x, on = "<joincolumn>"]`

# Using `dcast`

*Super Great Casting*

`dcast` works like `reshape::cast`, but it can also do multi-variable casting.

```{r}
dcast(dt, fname~type+ccode, value.var = "fname")
```

`data.table` can also `melt`.

# `set` functions

Mutability can make a gigantic difference when concerned with performance. 

data.table offers some methods for setting column names and order that don't copy data.

* `setnames`
* `setorder`
* `setcolorder`

# `set` functions

Setting column names with `setnames`: 

```{r}
names(dt)
setnames(dt, c("fname", "lname"), c("firstname", "lastname"))
names(dt)
```

# `set` functions

Setting column order with `setcolorder`:

```{r}
names(dt)
setcolorder(dt, c("lastname", "firstname"))
names(dt)
```

# `set` functions

Setting row order with `setorder`:

```{r}
setorder(dt, uid, date)
head(dt[,c(1:6)])
```

Know that there are two versions of this, one that accepts strings and
one that looks for variables in the `data.table` namespace.

See `?setorder` for more details.

# `setKey` and indexing

On thing that can make `data.table` very fast is the use of indexes
and keys.

* performance boost for large `data.table` join or subset operations
* no need to define `on` arguments when merging

# `setKey` and indexing

```{r}
uids <- fread("./data/slides/usertab.csv")
conn <- fread("./data/slides/conntab.csv")

setkey(uids, uid)
setkey(conn, uid)

head(uids[conn])
```

# Putting some of this stuff together...

What does this do?

```{r eval=F}
dt[lname == "Woods", note:="Don't talk about witches"]
```

# Assigning IDs by group...

```{r eval=F}
dt[, conn:=(1:.N), by = .(uid, type, date)]
```

# Grouping and chaining

```{r eval=F}
dt[, firstCon := date == min(date) & con == 1, by = uid
   ][, firstConIsDesk := firstCon & device == "desktop"
     ][ firstConIsDesk == 1 ]
```

# 

Lets work with some data!
